Automatically proving equivalence by type-safe reflection

### 2016-01-18 paper-JFP-DTP/content/2-basicIdeas.tex

When trying to prove this kind of equalities, the variables are abstracted, and they become part of the context. On the example 1 just above, after abstraction of the variables, the goal becomes simply $(x1 + x2) + (x3 + x4) = (x1 + (x2 + x3)) + x4$, which is something of the general form $x=y$.
The general idea --that will also apply for the more general problem detailed in the next sections-- will be to normalize both sides of the "potential equality" $x=y$, and afterwards to compare them using Leibniz syntactical equality.
The goal of the normalization is to compute a canonical representation for a number $x$, such that any other number provably equal to $x$ (by using the two available properties) will have the same canonical representation. For example, the normalisation might transform $x+((y+Z)+z)$ into $(x+y)+z$ if we decide that the normal form will be completely left associative, and that every addition between an element $a$ and zero should be simplified to $a$. It will then be possible to decide the equality by simply comparing the normalised left and right hand sides with a simple and strict syntactical equality. This is in fact what we do all the time when we have to decide if two things, written differently, are equal or not. When a human is given two mathematical polynomials and has to decide the equality of these functions, a technique that always works is to decide once and for all a canonical representation of polynomials, and to put both polynomials in this form. If the normalised forms are the same, then the two original polynomials are equal, otherwise they do not represent the same computation.\\
\\
In fact, such a normalisation function can't be written directly, because in the LHS and RHS of $x=y$, we potentially have variables which have been universally quantified. And the normalization function needs to do different treatments for a "variable natural number" (a number which has been universally quantified) and for the constant $Z$. This is not possible yet, because once the variables are abstracted, they are just ordinary values of type $Nat$, and there is no way to distinguish them. Indeed, this information only exists at the level of the ASTs representing the two terms, and we do not have a direct access to these syntax trees.
For this reason, we will work by reflection. In this little example, it means that we will define a data type which will be used as an encoding of natural numbers, or more precisely, an encoding of natural numbers composed of "variable numbers", $Z$, and additions of these things. This data type will allow us to know the internal structure of a number, ie, where the variables and constants are.
Indeed, this data type will allow pattern matching. Previously, we were only able to pattern match a natural number against the constructors $Z$ and $S$, which wasn't what we needed.  With the first approximation of the data type $Expr$ presented in the Fig 6, we will be allowed to pattern match an encoding of number against the constructor $Plus$, $Var$ and $Zero$, which gives us the informations we want.

@@

If we continue with this first definition of $Expr$, the normalisation function will certainly take an Expr and produce another Expr, and we will need to prove the following lemma afterwards : \\
$\forall\ e:Expr\ \Gamma,\ reify\ (reduce\ e)\ =\ reify\ e$ \\
where $reify$ is a function computing the interpretation of an Expr in a context $\Gamma$, that is to say, the natural number that this Expr is encoding.
This proof can be quite tricky to build because it relies on the complete behaviour of the reduction and on the way the interpretation is computed.
For this little example, the reduction procedure will not be too heavy, but in the next sections, with more sophisticated algebraic structures, we will have more rewriting rules to deal with and it will certainly become more problematic to "unfold" the definition of a gigantic reduction procedure. We want to avoid pain as much as possible, and for this reason, we don't want to have to write huge proofs. \\
\\
To avoid this source of complexity, we add an index to the type $Expr$, and this index is precisely the concrete number that the Expr is encoding. This is the first brick to our type-safe reflection mechanism. Thus, it won't be necessary to define the $reify$ function, as we will know directly the concrete element reflected by a term of type $Expr\ \Gamma\ x$ just by looking at its index $x$. Something more important is that we will also have a guarantee that the reflection of an expression $e$ is indeed a faithful representation of $e$ \\

>>>>

When trying to prove this kind of equalities, the variables are abstracted, and they become part of the context. On the example 1 just above, after abstraction of the variables, the goal becomes simply $(x1 + x2) + (x3 + x4) = (x1 + (x2 + x3)) + x4$, which is something of the general form $x=y$.
The general idea --that will also apply for the more general problem detailed in the next sections-- will be to normalize both sides of the "potential equality" $x=y$, and afterwards to compare them using Leibniz syntactical equality.
The goal of the normalization is to compute a canonical representation for a number $x$, such that any other number provably equal to $x$ (by using the two available properties) will have the same canonical representation. For example, the normalisation might transform $x+((y+Z)+z)$ into $(x+y)+z$ if we decide that the normal form will be completely left associative, and that every addition between an element $a$ and zero should be simplified to $a$. It will then be possible to decide the equality by simply comparing the normalised left and right hand sides with a simple and strict syntactical equality. This is in fact what we do all the time when we have to decide if two things, written differently, are equal or not. When a human is given two mathematical polynomials and has to decide the equality of these functions, a technique that always works is to decide once and for all a canonical representation of polynomials, and to put both polynomials in this form. If the normalised forms are the same, then the two original polynomials are equal, otherwise they do not represent the same computation.\\
\\
In fact, such a normalisation function can't be written directly, because in the LHS and RHS of $x=y$, we potentially have variables which have been universally quantified. And the normalization function needs to do different treatments for a "variable natural number" (a number which has been universally quantified) and for the constant $Z$. This is not possible yet, because once the variables are abstracted, they are just ordinary values of type $Nat$, and there is no way to distinguish them. Indeed, this information only exists at the level of the ASTs representing the two terms, and we do not have a direct access to these syntax trees.
For this reason, we will work by reflection. In this little example, it means that we will define a datatype that will be used as an encoding of natural numbers, or more precisely, an encoding of natural numbers composed of "variable numbers", $Z$, and additions of these things. This datatype will let us inspect the internal structure of a number by pattern matching.
Previously, we were only able to pattern match a natural number against the constructors $Z$ and $S$, which wasn't what we needed.  With the first approximation of the data type $Expr$ presented in the Fig 6, we will be allowed to pattern match an encoding of number against the constructors $Plus$, $Var$ and $Zero$.

@@

If we continue with this first definition of $Expr$, the normalisation function will certainly take an Expr and produce another Expr, and we will need to prove the following correctness lemma afterwards : \\
$\forall\ e:Expr\ \Gamma,\ reify\ (reduce\ e)\ =\ reify\ e$ \\
where $reify$ is a function computing the interpretation of an Expr in a context $\Gamma$, that is to say, the natural number that this Expr is encoding.
This proof can be quite tricky to build because it relies on the complete behaviour of the reduction and on the way the interpretation is computed.
For this little example, the reduction procedure won't be too heavy, but in the next sections, with more sophisticated algebraic structures, we will have more properties to deal with and it will certainly become more problematic to unfold the definition of a gigantic reduction procedure. We want to avoid pain as much as possible, and for this reason, we don't want to have to write big proofs, and even to avoid the writing of proofs as much as possible. \\
\\
To avoid this source of complexity, we add an index to the type $Expr$, and this index is precisely the concrete number that the Expr is encoding. This is the first brick to our type-safe reflection mechanism. Thus, it won't be necessary to define the $reify$ function, as we will know directly the concrete element reflected by a term of type $Expr\ \Gamma\ x$ just by looking at its index $x$. We also get the guarantee that the reflection of an expression $e$ is indeed a faithful representation of $e$ \\

### 2016-01-18 paper-JFP-DTP/content/6-discussion.tex

In Coq's latest implementation of a Ring solver, described in [1], they almost followed what has just been desribed as a "traditional" approach with the use of many auxiliary lemmas. The few differences essentially lie in the way the function $reflect$ is defined. In Coq, they use the language LTac to program this automatic reflection\footnote{This function --that they call mkPolexpr-- can be found in $plugins/ring/Ring\_tac.v$}. Ltac is an untyped tactic language. It is untyped in the sense that an Ltac "function" produces something which might not have a valid  --and unique-- type. Because of this, Ltac definitions can only be used in the context of goals. Thus, applying a tactic defined with LTac might work, and then it makes progress to the current goal, or it might fail, and in this case the goal is kept unchanged. And anyway, the safety of the proof done is going to be checked during the final QED. So, Ltac functions can't be used in the statement of a lemma, and it is therefore not possible to reason about them. That means that it is not possible to write the lemma $reflect\_and\_reify\_correct$ of the "traditional" approach described in the previous subsection, because it is even impossible to state it as it uses a function defined in LTac. Because they can't even state this property, they can't finish the proof of the main theorem $decideEq\_correct\ :\ \forall\  x\ y:c,\ decideEq\ x\ y\ =\ true\ \rightarrow\ x \simeq\ y$. Indeed, the last step of this proof was to apply two times --one for the LHS, one for the RHS-- the fact that $\forall\ x:c,\ reify\ (reflect\ x)\ \simeq\ x$. One possibility would have been to add this axiom, but this is particularly unsightly, and potentially harmful. Also, it would imply that anyone using the ring prover would have this axiom added to his development, and would be forced to believe in it. Some proofs would be done automatically for the user of the system, but at the overly-expensive price of adding some uncertainty.

>>>>

In Coq's latest implementation of a Ring solver, described in~\cite{Coq2005}, they almost followed what has just been desribed as a "traditional" approach with the use of many auxiliary lemmas. The few differences essentially lie in the way the function $reflect$ is defined. In Coq, they use the language LTac to program this automatic reflection\footnote{This function --that they call mkPolexpr-- can be found in $plugins/ring/Ring\_tac.v$}. Ltac is an untyped tactic language. It is untyped in the sense that an Ltac "function" produces something which might not have a valid  --and unique-- type. Because of this, Ltac definitions can only be used in the context of goals. Thus, applying a tactic defined with LTac might work, and then it makes progress to the current goal, or it might fail, and in this case the goal is kept unchanged. And anyway, the safety of the proof done is going to be checked during the final QED. So, Ltac functions can't be used in the statement of a lemma, and it is therefore not possible to reason about them. That means that it is not possible to write the lemma $reflect\_and\_reify\_correct$ of the "traditional" approach described in the previous subsection, because it is even impossible to state it as it uses a function defined in LTac. Because they can't even state this property, they can't finish the proof of the main theorem $decideEq\_correct\ :\ \forall\  x\ y:c,\ decideEq\ x\ y\ =\ true\ \rightarrow\ x \simeq\ y$. Indeed, the last step of this proof was to apply two times --one for the LHS, one for the RHS-- the fact that $\forall\ x:c,\ reify\ (reflect\ x)\ \simeq\ x$. One possibility would have been to add this axiom, but this is particularly unsightly, and potentially harmful. Also, it would imply that anyone using the ring prover would have this axiom added to his development, and would be forced to believe in it. Some proofs would be done automatically for the user of the system, but at the overly-expensive price of adding some uncertainty.

### 2016-01-18 paper-JFP-DTP/paper.tex
Check
paper : use of listings and corrections
Added new section

\usepackage{float}

@@

\input{./content/5-discussion.tex}

\begin{thebibliography}{}
 \bibitem[\protect\citename{B. Gregoire and A. Mahboubi}2005]{Coq2005}
   Gregoire,~B. and Mahboubi,~A. (2005) Proving equalities in a commutative ring done right in Coq, TPHOLs 2005, Oxford, UK, August 22-25, 2005, Proceedings

 \bibitem[\protect\citename{W.A Howard}1980]{How80}
   Howard W.A. The formulae-as-types notion of construction. In J. R. Seldin and J. P. Hindley, editors, To H. B. Curry: Essays on Combinaory Logic, Lambda Calculus, and Formalism. Academic Press, 1980.

\end{thebibliography}

>>>>

\usepackage{float}
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily,
  columns=fullflexible,
  keepspaces=true,
  mathescape
}

@@

\input{./content/5-results.tex}

\input{./content/6-discussion.tex}

\bibliographystyle{jfp}
\bibliography{dtp}

%\begin{thebibliography}{}
% \bibitem[\protect\citename{B. Gregoire and A. Mahboubi}2005]{Coq2005}
%   Gregoire,~B. and Mahboubi,~A. (2005) Proving equalities in a commutative ring done right in Coq, TPHOLs 2005, Oxford, UK, August 22-25, 2005, Proceedings

% \bibitem[\protect\citename{W.A Howard}1980]{How80}
%   Howard W.A. The formulae-as-types notion of construction. In J. R. Seldin and J. P. Hindley, editors, To H. B. Curry: Essays on Combinaory Logic, Lambda Calculus, and Formalism. Academic Press, 1980.

%\end{thebibliography}
